Staged Abstract Interpreters

### 2019-02-11 paper/10_intro.tex
revise

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, and then
evaluates the static part, and generates a residual program that relies on the
dynamic part. However, precisely analyzing binding-time for an arbitrary program
is hard in general. As an alternative and pragmatic approach to specialization
and partial evaluation, multi-stage programming (MSP for short)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage annotations
(i.e., binding-time annotations) from the programmers. These staging annotations
identify which part of the input program should be specialized. As a classical
example, we consider a power function:

@@

If the program identifies that @b@ will be known dynamically (as shown in its
type @Rep[Int]@ -- a representation of @Int@) and @e@ is statically known, say
5, then we can specialize the function \texttt{power} with respect to \texttt{e
= 5} and generate a new specialized function that runs faster:

@@

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach
to construct sound static analysis by approximation is almost as old as Futamura
projections. Some recent progress such as Abstracting Abstract Machines (AAM)
uncovers a methodology to derive sound abstract interpreters from their concrete
counterparts and has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}. Given the structural
similarity between concrete and abstract interpreters, intellectually it is
natural to raise the question whether it is possible to specialize an abstract
interpreter, and how can we effectively do that having considered their
different functionalities. This paper studies the confluence of two old ideas
--- Futamura projection and abstract interpretation, but from the perspective of
their recent realizations --- multi-stage programming and definitional abstract
interpreters. To be specific, we present the application of the first Futamura
projection on abstract interpreters, and the approach of constructing optimizing
abstract interpreters by multi-stage programming.

@@

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, and then
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-time is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage
annotations (i.e., binding-time annotations) to be explict from the programmers.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we consider the power function:

@@

If the programmer identifies that @b@ will be known in the future stage (as shown
in its type @Rep[Int]@ -- a representation of @Int@) and @e@ is known in the
current stage, say 5, then we can specialize the function @power@ with
respect to @e = 5@ and generate a specialized function that runs faster:

@@

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fixpoints \cite{DBLP:conf/popl/CousotC77}.
Some recent progress such as Abstracting Abstract Machines (AAM)
uncovers a methodology to derive sound abstract interpreters from their concrete
counterparts and has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}. Given the structural
similarity between concrete and abstract interpreters, intellectually it is
natural to raise the question whether it is possible to specialize an abstract
interpreter, and how can we effectively do that having considered their
different functionalities. This paper studies the confluence of two old ideas
--- Futamura projection and abstract interpretation, but from the perspective of
their recent realizations --- multi-stage programming and definitional abstract
interpreters. To be specific, we present the application of the first Futamura
projection on abstract interpreters, and the approach of constructing optimizing
abstract interpreters by multi-stage programming.

@@

\note{
  Expression problem \cite{DBLP:conf/icfp/GibbonsW14, DBLP:conf/ecoop/KrishnamurthiFF98, expproblem},
  deep embedding, four different interpretations.
}

### 2019-02-11 paper/paper.tex
reorganize and address typos
revert to review mode
revise

\documentclass[sigplan,nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
% \documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
%\acmConference[Submitted to PLDI'18]{ACM SIGPLAN Conference on Programming Language Design and Implementation}{June 18--22, 2018}{Philadelphia, PA, USA}
\acmConference[Submitted to PLDI]{}{June 2019}{Phoenix, Arizona}
\acmYear{2019}
%\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

@@

@@

>>>>

%\documentclass[sigplan,nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2019}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

@@

\iffalse

@@

\fi
